Diversity, Segregation, and Information in Cities

### 2016-11-10 submission.tex

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{I}nformation theory...blah blah blah.

\section*{Results}
	We model a city as a set of individuals distributed over a continuous spatial region $R$. 
	Each individual can belong to one of a set $\mathcal{Y}$ of racial groups, such as $\mathcal{Y} = \{\text{Asian}, \text{ Black}, \text{ Hispanic}, \text{ Other}, \text{ White}\}$. 
	The cross-tabulation of individuals and groups defines a probability $p(X,Y)$ that a randomly chosen individual from the city will live at location $X \in R$ and belong to group $Y$. 
	The population density at $x$ is $p(x) = \sum_{y \in \mathcal{Y}} p(x,y)$. 
	The global proportion of residents belonging to group $y$ is $p(y) = \int_R p(x,y)\; dx$. 
	The proportion of residents at $x$ belonging to group $y$ is $p(y|x) = p(x,y) / p(x)$. 
	To compare distributions to one another we use the Kullback-Leibler divergence 
	\begin{equation}
		D[q(A)\|r(A)] = \sum_{a \in \mathcal{A}} q(a) \log \frac{q(a)}{r(a)}\;.
	\end{equation}
	The divergence $D$ measures dissimilarity between distributions: it is zero if and only if $q = r$, and grows larger when $q$ and $r$ assign greater probability to different events.

\subsection*{Entropy and Global Diversity}
	\emph{Global} measures of the city depend only on $p(Y)$, the global composition of racial groups. 
	The global Shannon entropy $H(Y)$ measures global diversity by comparing the global distribution $p(Y)$ to the uniform distribution $u(Y)$: 
	\begin{equation}
		H(Y) = -\sum_{y \in \mathcal{Y}}p(y) \log p(y) = \log \abs{\mathcal{Y}} - D[p(Y)\|u(Y)]\;.
	\end{equation}

	Local-to-global measures depend on comparisons between properties encoded by the local compositions $p(y|x)$ and the global composition $p(y)$.

	Local measures depend only on comparisons between local compositions $p(y|x)$ and other nearby local compositions, without reference to the global composition $p(y)$. 
\begin{figure}%[tbhp]
		\centering % need to center this
		\includegraphics[width=\linewidth]{figs/checkerboard.pdf}
		\begin{tabular}{l | c c c} 
		City & $H(Y)$ & $I(X,Y)$ & $J(X,Y)$ \\
		\hline
		(a) & $0$        & $0$        & $0$\\
		(b) & $\log 2$ & $0$        & $0$\\
		(c) & $\log 2$ & $\log 2$ & $\frac{1}{7} \log 2$\\
		(d) & $\log 2$ & $\log 2$ & $\log 2$
		\end{tabular}
		\caption{
			\todo[inline=true]{Illustrate computation of $J$ on this fig?}
			Schematic cities illustrating three dimensions of segregation. 
			City (a) lacks global diversity, while cities (b)-(d) are maximally globally diverse. 
			City (b) is perfectly spatially even, whereas cities (c) and (d) are maximally uneven. 
			City (d) has higher spatial exposure than city (c) due to more contact points between racially distinct neighborhoods.
		} \label{fig:checkerboard}
	\end{figure}
\subsection*{Mutual Information and Spatial Evenness}
	Local-to-global measures depend on comparisons between properties encoded by the local compositions $p(Y|x)$ and the global composition $p(Y)$. 
	When each of the compositions $p(Y|x)$ is the same as $p(Y)$, the city is perfectly spatially even -- every racial group is represented at each location in proportion to their presence in the city overall. 
	When the local compositions $p(Y|x)$ differ from $p(Y)$, the city is spatially uneven. 
	We can measure spatial evenness as the average difference between local and global compositions: 
	\begin{equation}
		I(X,Y) = \E_X[D[p(Y|X)\|p(Y)]\;.
	\end{equation}
	The operator $\E_X$ denotes average or expectation with respect to the probability distribution of $X$. 
	$I(X,Y)$ is the \emph{mutual information}, a fixture of information theory. \cite{Bettencourt2015}

	When all groups are evenly distributed, such as in Figure \ref{fig:checkerboard}(b), each location has the same composition as the global composition, and $I(X,Y) = 0$. 
	On the other hand, $I(X,Y)$ achieves its maximum value of $\log \abs{\mathcal{Y}}$ when every location is monogroup, such as in Figure \ref{fig:checkerboard}(c).

@@

	Global and local-global measures can illuminate important high-level urban structure, but miss an important class of phenomena. 
	The model two-race cities (c) and (d) in Figure \ref{fig:checkerboard} each have the same values of $H(Y)$ and $I(X,Y)$. 
	City (c), however, has a single dividing line of racial separation, while city (d) has smaller pockets of racial distinctiveness. 
	In sociology or urban planning, this may be the difference between metastasized segregation and healthy variation. 
	The difference is intrinsically local, defined in terms of relations between local compositions $p(Y|X)$ and nearby other local compositions.

	To measure local variation, imagine segmenting the city into a grid of small neighborhoods. 
	The \emph{local mutual information} in the neighborhood $B$ is the mutual information between $X$ and $Y$, restricted to $B$:
	\begin{equation*}
		I(X,Y | X \in B) = \E_{X|X \in B}[D[p(Y|X,X\in B)\|p(Y|X\in B)]\;.
	\end{equation*}
	The \emph{mean local information} measures the average local variation in group composition across the entire city:  
	\begin{equation}
		J(X,Y) = \E_B[I(X,Y|X \in B)]\;.
	\end{equation}
	If we regard the data as a discretization of a smooth probability field, then it is possible to show using arguments from information geometry \cite{Amari2000} that the local information in a neighborhood $B$ of radius $r$ is an approximation of the Fisher information $\mathcal{J}$:
	\begin{equation}
		I(X,Y|X \in B) \cong \frac{r^2}{4} \text{tr } \mathcal{J}_Y(x)\;, \label{eq:fisher_approx}
	\end{equation}
	where $[\mathcal{J}_Y(x)]_{ij} = \E_Y[-\partial_{x_i} \partial_{x_j} \log p(Y|x)]$ is the Fisher information matrix and $\text{tr } \mathcal{J}_Y(x)$ is the sum of its eigenvalues.

@@

			\caption{
				Spatial analysis of residential segregation in Atlanta and Philadelphia. 
				(\emph{Top}): Percentage of black residents. 
				(\emph{Middle}): Hexgrid of spatial resolution 1km imposed over each city. Within each hex $B$ we compute the local information local information $J_Y(B) = I(X,Y | X \in B)$. 
				(\emph{Bottom}): Numerical summary, including the three information measures.
			} \label{fig:Atlanta_philly}
	\end{figure}

\subsection*{Comparative Urban Analysis}

@@

	\begin{figure}
		\includegraphics[width=\linewidth]{figs/mutual_fisher.pdf}
		\caption{
			Comparative global segregation measures in major U.S. cities, measured using racial categories $\mathcal{Y} = \{\text{Asian}, \text{ Black}, \text{ Hispanic}, \text{ Other}, \text{ White}\}$. 
			We used a hexgrid of spatial resolution $r = 1$km for the computation of $J(X,Y)$.
		} \label{fig:mutual_fisher}
	\end{figure}

\subsection*{Identifying Sociospatial Structure} 
	The measures $H(Y)$, $I(X,Y)$, and $J(X,Y)$ quantify the presence of global, local-global, and purely local sociospatial structure. We can also use information methods to identify and visualize that structure directly. While approaches exist to this problem,\footnote{For three, see \cite{Logan2011}} these approaches tend to conceptualize this structure in terms of monoracial subregions. In practice, however, there may be coherent biracial or multiracial patterns in space. Fortunately, it is possible to motivate a structure-finding algorithm for identifying high-level demographic trends without restriction to monoracial regions.

	\begin{figure}

		\includegraphics[width = .5\linewidth]{figs/Atlanta_clusters_binary_6.pdf} 
		\includegraphics[width = .5\linewidth]{figs/Philadelphia_clusters_binary_6.pdf} \\
		\includegraphics[width = .5\linewidth]{figs/Atlanta_clusters_binary_20.pdf}
		\includegraphics[width = .5\linewidth]{figs/Philadelphia_clusters_binary_20.pdf} \\

		\centering
		\begin{tabular}{l | c c c}
			\bfseries City  & $I(C_6, Y)$ & $I(C_{20},Y)$ & $I(X,Y)$ \\\hline
			\csvreader[late after line=\\]{figs/clustering_summary.csv}{}
			{\csvcoli & \csvcoliii & \csvcoliv & \csvcolii}
		\end{tabular}
		\caption{
			Example clusterings in Atlanta and Philadelphia using \eqref{eq:cluster_opt}, using $6$ and $20$ clusters. 
			The table below summarizes the between-groups mutual information captured by each cluster and compares it to the mutual information in the unaggregated blockgroup level data.
		} \label{fig:clusterings}

	\end{figure}
	Let $C$ be a random variable with the property that $H(C|X) = 0$; i.e. $X$ completely determines $C$. We can think of $C$ as group labels for the spatial locations $X$. Then, the chain rule of information\footnote{See \cite{Cover1991}} implies that 
	\begin{equation}
		I(X,Y) = \underbrace{I(C,Y)}_{\text{between-group}} + \underbrace{I(X,Y|C)}_{\text{within-group}}. \label{eq:chain_rule}
	\end{equation}

	Equation \eqref{eq:chain_rule} motivates a natural algorithm for identifying sociospatial structure. The core idea is that a clustering $C$ is ``good'' when it captures most of the intrinsic information in the system in the between-groups term $I(C,Y)$. Mathematically, we therefore wish to maximize $I(C,Y)$ subject to a fixed number of clusters, an approach popularized by \cite{Dhillon2003,Banerjee2005}. In full generality, this is a challenging discrete optimization problem that may not be computationally tractable. Additionally, for interpretability, we need to apply a spatial constraint: locales should be clustered together only when they are adjacent in space. To handle these two problems, we use a greedy algorithm based on agglomerative clustering. At stage of the algorithm, we consider the problem of choosing a pair of locales $\{i*, j*\}$ to aggregate together. A good choice of locales leads to minimal information loss, which we may quantify as follows. The mutual information over the full complement $\mathcal{K}$ of locales may be written: 
	\begin{align*}
	 	I(X,Y) &= \sum_{k \in \mathcal{K}} p(X = k) D[p(Y|X = k\|p(Y)] 
	\end{align*} 
	The contribution of locale $k$ to this sum is 
	\begin{align*}
		I_{k}(X,Y) &= p(X = k)D[p(Y|X = k)\|p(Y)]. 
	\end{align*}
	If we replace $i$ and $j$ with an aggregated locale $i\cup j$, the aggregated locale's contribution to the mutual information is 
	\begin{align*}
		I_{i\cup j}(X,Y) = p(X \in \{i,j\}) D[p(Y|X \in \{i,j\}) \| p(Y)]\;.
	\end{align*}
	The loss associated with aggregating $i$ and $j$ is the difference between their individual contributions and the aggregated contribution of $i\cup j$:
	\begin{align*}
		d(i,j) = I_{i}(X,Y) + I_{j}(X,Y) - I_{i \cup j}(X,Y).  
	\end{align*}
	The function $d$ may be regarded as a similarity measure. It satisfies $d(i,j) = 0$ if and only if $p(Y|X = i) = p(Y|X = j)$; i.e. the demographic profiles of locations $i$ and $j$ are identical. It is also symmetric in $i$ and $j$. It is not a true metric because it does not satisfy the triangle inequality. It can nevertheless be used to motivate a form of agglomerative hierarchical clustering, where at each phase of the algorithm we compute 
	\begin{equation}
		(i^*, j^*) = \argmin_{i \text{ neighbors } j} \; d(i,j)\; \label{eq:cluster_opt}
	\end{equation}
	and combine $i^*$ and $j^*$ into a single new region. We repeat this procedure until only $N$ regions remain. The optimization over $i$ and $j$ ensures that the groupings so defined approximate maximum between-group information partitions, and the constraint that $i$ must neighbor $j$ ensures that the structure so determined is spatially contiguous. Equation \eqref{eq:cluster_opt} thus defines \emph{spatially constrained, information-maximizing agglomerative clustering}. Since the algorithm is greedy, it possesses no guarantees of optimality, but in practice its performance leads to intuitive, demographically-coherent regions. 
	\todo[inline=true]{Should we state the algorithm in a figure?}

	\begin{figure}
		\centering
			% \begin{subfigure}[b]{0.45\linewidth}
			\includegraphics[width = .4\linewidth]{figs/example_cluster_map.pdf}
			% \caption{} \label{subfig:detroit_map}
			% \end{subfigure}
			% \begin{subfigure}[b]{0.45\linewidth}
			\includegraphics[width = .4\linewidth]{figs/example_clusters_detailed.pdf}
			% \caption{} \label{subfig:detroit_groups}
			% \end{subfigure}
			\caption{
				Multigroup sociospatial structure in Detroit. 
				\emph{(Left)}: Map of six clusters. 
				\emph{(Right)}: Joint distribution of demographics and clusters. 
				Shadings reflect population proportions within clusters. 
				Numerical labels give raw population proportions; for example, 29\% of the population is Black and resides in cluster B. 
				Detroit has $H(Y) = 1.12$, $I(X,Y) = 0.57$, and $J(X,Y) = 0.08$. 
				This clustering captures $I(C_6,Y) = 0.36$ nats of information.
			} \label{fig:detroit}
	\end{figure}

	This technique extends naturally to multi-group settings, though the visualization of results requires more subtlety. In Figure \ref{fig:detroit}, we conduct an illustrative analysis of multigroup segregation segregation in Detroit and the surrounding Wayne County. The prominent clusters A and B reflect the primary white/black division in the city, with A representing the white suburbs and B the denser black inner city. Cluster C is interpretable as a transitional area between these two populations, with both white and black residents represented. Cluster D coincides with the historically distinct, predominantly white town of Grosse Pointe. Cluster E coincides with the city of Hamtramck, whose history has been shaped by high levels of immigration from Europe and Asia. Finally, Cluster F coincides with the Hispanic neighborhood Mexicantown.

	Aside from simply highlighting geographic trends, this analysis also enables the quantitative study of traditional concerns in segregation studies, such as concentration. According to \cite{Massey1988}, ``\emph{concentration} refers to the degree of a group's agglomeration in urban space.'' The concept of concentration invites us to consider the composition and density of each cluster. For example, whites make up 70\% of ``their'' cluster A, while black residents make up 86\% of ``their'' cluster B. Cluster B is also more densely populated than Cluster A: it contains 34\% of Wayne County's population, but just 19\% of the analyzed geographic area, making it more than twice as dense as Cluster A. Thus, in and around Detroit, black neighborhoods tend to be both more racially homogeneous and more densely packed in urban space than white ones. We can also learn about pairwise group exposure. For example, from Figure  \ref{subfig:detroit_groups}, we observe that Asian residents tend to be found alongside white residents, but more rarely alongside black or Hispanic residents. Hispanic residents tend to group in Mexicantown and in predominantly white neighborhoods -- not black ones.

\subsection*{Evaluating Clusterings}
	As discussed, the mean local information $J(X,Y)$ can be viewed as a measure of the number and character of transitional areas between regions of differing demographic structure.  It is low in Atlanta because the city is dominated by one, large North-South boundary while it is higher in Philadelphia due to many points of contact between neighborhoods of differing composition. Sociospatial transitions are also precisely the ``fault lines'' along which we would expect the clustering algorithm \eqref{eq:cluster_opt} to split the city. It is therefore not a coincidence that Philadelphia both has a higher value of $J(X,Y)$ and requires a greater number of clusters to convey a similar amount of sociospatial information.

>>>>

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{T}his is a story about the hierarchical structure of demographic and spatial form. We need to tell this story because traditional methods only give us a single number, whereas what we really want to see is spatial structure at varying levels of aggregation. This is a story about practical computation for seeing the hierarchical structure of social variation. This is a story about finding the smallest unit of spatial variation -- the local information -- and then scaling up in a smart way. It's a story about quantifying the intrinsic structure and then going into the world of algorithms. It's a story about computing with the hierarchical structure. It's a story about explicitly incorporating space on different levels. It's a story about not-too-novel algorithms applied in new ways. It's a story about finding a principled role for information theory in the study of urban form. It's a story about mathematical coherence. It's a story about love. It's a story about non-convexity and choosing to be ok with that. It's a story about finding new ways to view what's already visible, to organize what's there in a useful way. It's a story about computation and open data. It's open science. It's a story about not everything needing to be statistics. It's a story about taking the data as it is. It's a story about how to learn from the Census. It's a story about how we can improve on sociological methodology using results from physics and mathematics. It's a story about how physics and mathematics are missing insights that you only get from the social sciences. It's a story about collaboration. It's a story about how one day we'll get the dynamical picture too but we haven't coded that up yet or understood the mathematics. It's not yet a story about the quadratic structure of local variation through derived dispersions, but we'll get there one day even if that's just on arXiv. It's a story about the hierarchical structure, about different ways that information scales in different places.

Multiple scales. How can we learn to see them all? Well, we can consider properties of information scaling at various levels. But it won't really do to organize these levels in arbitrary ways, we should actually seek the structure as best we can. Not optimally, but as best we can. Try to see what's already there, what's in front of us. Nothing is hidden. We can do some motivation around fractical structure, maybe cite something from batty or PNAS on fractal structure of cities and fractal structure of demographics, little pockets of variability. Just look at the US, where you have both global trends across the whole country and then smaller trends within each area and then within each neighborhood. Maybe New York is an example, there are different little pockets within each borough as well as trends between boroughs. So we can tell a bit of a story that way. Recall Roberto's question: is Detroit segregated?

The agenda is to flexibly study demographic separation in space on varying scales. To do that we first need to understand demographic separation on the lowest level, and this requires calculus. So we do the calculus and we find that we get the trace of the Fisher information out at the end, which is pleasing. Ok, so that's the local picture, hooray! One way to go from the local picture to the global picture is just to integrate over everything, after which we find that we get a nice averaged measure that captures the high-level differences between cities. But this is in some sense a blunt instrument; we don't exactly learn too much about in what ways they are different. A more subtle instrument is to use an algorithm -- the information-maximizing agglomerative clustering -- to find the structure itself, and then we can quantify the performance of the algorithm and take that as a measure of complexity directly. What's nice about this is that our complexity measure is actually directly related to our ability to efficiently summarise the structure of urban form. Perhaps we should state the algorithm explicitly as a figure. Our ultimate goal is to show that this material is of relevance to a fairly broad audience, hopefully including some sociologists, and we can do some evangelizing at conferences etc.

\section*{Results}
	We model a city as a set of individuals distributed over a continuous spatial region $R$. 
	Each individual can belong to one of a set $\mathcal{Y}$ of racial groups, such as $\mathcal{Y} = \{\text{Asian}, \text{ Black}, \text{ Hispanic}, \text{ Other}, \text{ White}\}$. 
	The cross-tabulation of individuals and groups defines a probability $p(X,Y)$ that a randomly chosen individual from the city will live at location $X \in R$ and belong to group $Y \in \mathcal{Y}$. 
	The population density at $x$ is $p(x) = \sum_{y \in \mathcal{Y}} p(x,y)$. 
	The global proportion of residents belonging to group $y$ is $p(y) = \int_R p(x,y)\; dx$. 
	The proportion of residents at $x$ belonging to group $y$ is $p(y|x) = p(x,y) / p(x)$. 
	To compare distributions to one another we use the Kullback-Leibler divergence 
	\begin{equation}
		D[q(A)\|r(A)] = \sum_{a \in \mathcal{A}} q(a) \log \frac{q(a)}{r(a)}\;.
	\end{equation}
	The divergence $D$ measures dissimilarity between distributions: it is zero if and only if $q = r$, and grows larger when $q$ and $r$ assign greater probability to different events.

\subsection*{Entropy and Global Diversity}
	A global measure depends only on $p(Y)$, the global composition of racial groups. 
	The global Shannon entropy $H(Y)$ measures global diversity by comparing the global distribution $p(Y)$ to the uniform distribution $u(Y)$: 
	\begin{equation}
		H(Y) = -\sum_{y \in \mathcal{Y}}p(y) \log p(y) = \log \abs{\mathcal{Y}} - D[p(Y)\|u(Y)]\;.
	\end{equation}
	Since $H(Y)$ is a global metric, it gives a high-level summary but is insensitive to spatial structure.

\begin{figure}%[tbhp]
		\centering % need to center this
		\includegraphics[width=\linewidth]{figs/checkerboard.pdf}
		\begin{tabular}{l | c c c} 
		\textbf{City} & $H(Y)$ & $I(X,Y)$ & $J(X,Y)$ \\
		\hline
		(a) & $0$        & $0$        & $0$\\
		(b) & $\log 2$ & $0$        & $0$\\
		(c) & $\log 2$ & $\log 2$ & $\frac{1}{7} \log 2$\\
		(d) & $\log 2$ & $\log 2$ & $\log 2$
		\end{tabular}
		\caption{
			\emph{(Top)}: Schematic cities illustrating three dimensions of segregation. 
			City (a) lacks global diversity, while cities (b)-(d) are maximally globally diverse. 
			City (b) is perfectly spatially even, whereas cities (c) and (d) are maximally uneven. 
			City (c) has lower spatial exposure than city (d), since a typical small area of (c) (shown in highlight) is monoracial.
			In contrast, a typical small neighborhood of (d) has multiple contact points between racially distinct areas. 
			\emph{(Bottom)}: Information measures of each city. Entropy separates cities (a) and (b); mutual information separates (b) and (c), and mean local information separates (c) and (d). 
		} \label{fig:checkerboard}
	\end{figure}
\subsection*{Mutual Information and Spatial Evenness}
	Local-to-global measures depend on comparisons between properties encoded by the local compositions $p(Y|x)$ and the global composition $p(Y)$. 
	When each of the compositions $p(Y|x)$ is the same as $p(Y)$, the city is perfectly spatially even -- every racial group is represented at each location in proportion to their presence in the city overall. 
	When the local compositions $p(Y|x)$ differ from $p(Y)$, the city is spatially uneven. 
	We can measure spatial evenness as the average difference between local and global compositions: 
	\begin{equation}
		I(X,Y) = \E_X[D[p(Y|X)\|p(Y)]\;.
	\end{equation}
	The operator $\E_X$ denotes average or expectation with respect to the probability distribution of $X$. 
	The quantity $I(X,Y)$ is called the \emph{mutual information} in information theory \cite{Cover1991,Csiszzr2004,Bettencourt2015}.  
	When all groups are evenly distributed, such as in Figure \ref{fig:checkerboard}(b), each location has the same composition as the global composition, and $I(X,Y) = 0$. 
	On the other hand, $I(X,Y)$ achieves its maximum value of $\log \abs{\mathcal{Y}}$ when every location is monogroup, such as in Figure \ref{fig:checkerboard}(c).

@@

	Global and local-to-global measures can illuminate important high-level urban structure, but miss an important class of phenomena. 
	The model two-race cities (c) and (d) in Figure \ref{fig:checkerboard} each have the same values of $H(Y)$ and $I(X,Y)$. 
	City (c), however, has a single dividing line of racial separation, while city (d) has smaller pockets of racial distinctiveness. 
	In sociology or urban planning, this may be the difference between metastasized segregation and healthy variation. 
	The difference is intrinsically local: while the typical small neighborhood of city (c) is monoracial, the typical small neighborhood of (d) contains multiple contact points between different racial groups.

	To measure local variation, imagine segmenting the city into a grid of small cells. 
	The \emph{local mutual information} in the cell $B$ is the mutual information between $X$ and $Y$, restricted to $B$:
	\begin{equation*}
		I(X,Y | X \in B) = \E_{X|X \in B}[D[p(Y|X,X\in B)\|p(Y|X\in B)]\;.
	\end{equation*}
	The local information can be interpreted as the degree of change in demographics one would observe if one started in $B$ and then walked a small distance away.

	If we regard the data as a discretization of a smooth probability field, then it is possible to show using arguments from information geometry \cite{Amari2000} that the local information in a cell $B$ of radius $r$ is an approximation of the Fisher information $\mathcal{J}$:
	\begin{equation}
		I(X,Y|X \in B) \cong \frac{r^2}{4} \text{tr } \mathcal{J}_Y(x)\;, \label{eq:fisher_approx}
	\end{equation}
	where $[\mathcal{J}_Y(x)]_{ij} = \E_Y[-\partial_{x_i} \partial_{x_j} \log p(Y|x)]$ is the Fisher information matrix and $\text{tr } \mathcal{J}_Y(x)$ is the sum of its eigenvalues.

	The \emph{mean local information} measures the average local variation in group composition across the entire city:  
	\begin{equation}
		J(X,Y) = \E_B[I(X,Y|X \in B)]\;.
	\end{equation}
	Figure \ref{fig:Atlanta_philly} shows the computation of $J(X,Y)$ for Atlanta and Philadelphia. While the two cities have similar entropies $H(Y)$ and mutual informations $I(X,Y)$, their spatial structures are substantially different. Atlanta is characterized by a stark North-South divide between white and black residents. As a result, there are many areas in Atlanta where the local information is near $0$. These are areas in which one can walk for $1km$ or more without experiencing significant demographic change. Such areas are relatively rare in Philadelphia, which consists of a finer patchwork of demographically varying neighborhoods. Philadelphia therefore has a higher value of mean local information, reflecting greater spatial complexity in demographic variation.

@@

			\caption{
				(\emph{Top}): Percentage of black residents. 
				(\emph{Middle}): Hexgrid of spatial resolution 1km imposed over each city. Within each hex $B$ we compute the local information local information $J_Y(B) = I(X,Y | X \in B)$. 
				(\emph{Bottom}): Numerical summary including entropy, mutual information, and mean local information. 
			} \label{fig:Atlanta_philly}
	\end{figure}

\subsection*{Separation Profiles in the US}

@@

	\begin{figure}
		\includegraphics[width=\linewidth]{figs/mutual_fisher.pdf}
		\caption{
			Comparative global segregation measures in major U.S. cities, measured using racial categories $\mathcal{Y} = \{\text{Asian}, \text{ Black}, \text{ Hispanic}, \text{ Other}, \text{ White}\}$. 
			A hexgrid of spatial resolution $r = 1$km was for the computation of $J(X,Y)$. The grey line is the linear least-squares regression line provided as a visual aid. 
		} \label{fig:mutual_fisher}
	\end{figure}

\subsection*{Identifying Sociospatial Structure} 
	The measures $H(Y)$, $I(X,Y)$, and $J(X,Y)$ quantify the presence of global, local-global, and purely local sociospatial structure. We can also use information methods to identify and visualize that structure directly. While approaches exist to this problem,\footnote{For three, see \cite{Logan2011}} these approaches tend to conceptualize this structure in terms of monoracial subregions. In practice, however, there may be coherent biracial or multiracial patterns in space. Information methods define a useful approach. 
	\begin{figure}

		\includegraphics[width = .5\linewidth]{figs/Atlanta_clusters_binary_4.pdf} 
		\includegraphics[width = .5\linewidth]{figs/Philadelphia_clusters_binary_4.pdf} \\
		\includegraphics[width = .5\linewidth]{figs/Atlanta_clusters_binary_20.pdf}
		\includegraphics[width = .5\linewidth]{figs/Philadelphia_clusters_binary_20.pdf} \\

		\centering
		\begin{tabular}{l | c c c}
			\bfseries City  & $I(C_4, Y)$ & $I(C_{20},Y)$ & $I(X,Y)$ \\\hline
			\csvreader[late after line=\\]{figs/clustering_summary.csv}{}
			{\csvcoli & \csvcoliii & \csvcoliv & \csvcolii}
		\end{tabular}
		\caption{
			\emph{(Top)}:Example clusterings in Atlanta and Philadelphia using \eqref{eq:cluster_opt}, using $4$ and $20$ clusters. 
			\emph{(Bottom)}: Comparison of information content of clusterings and total mutual information in each city. 
		} \label{fig:clusterings}
	\end{figure}
	Let $C$ be a grouping of the locations $X$. Mathematically, $C$ is a random variable with the property that that $H(C|X) = 0$. The chain rule of information\footnote{See \cite{Cover1991}} implies that 
	\begin{equation}
		I(X,Y) = \underbrace{I(C,Y)}_{\text{between-group}} + \underbrace{I(X,Y|C)}_{\text{within-group}}. \label{eq:chain_rule}
	\end{equation}
	Analogous to sum-of-squares maximization in statistics, we wish to maximize $I(C,Y)$ subject to a fixed number of clusters, an approach popularized by \cite{Dhillon2003,Banerjee2005}. In full generality, this is a challenging discrete optimization problem that may not be computationally tractable. Additionally, for interpretability, should be clustered together only when they are adjacent in space. To handle these two problems, we use a greedy algorithm based on agglomerative clustering. At stage of the algorithm, we consider the problem of choosing a pair of locales $\{i*, j*\}$ such that aggregating them minimizes information loss. We quantify the loss as follows. The mutual information over the full complement $\mathcal{K}$ of locales may be written: 
	\begin{align*}
	 	I(X,Y) &= \sum_{k \in \mathcal{K}} p(X = k) D[p(Y|X = k\|p(Y)] 
	\end{align*} 
	The contribution of locale $k$ to this sum is 
	\begin{align*}
		I_{k}(X,Y) &= p(X = k)D[p(Y|X = k)\|p(Y)]. 
	\end{align*}
	If we replace $i$ and $j$ with an aggregated locale $i\cup j$, the aggregated locale's contribution to the mutual information is 
	\begin{align*}
		I_{i\cup j}(X,Y) = p(X \in \{i,j\}) D[p(Y|X \in \{i,j\}) \| p(Y)]\;.
	\end{align*}
	The loss associated with aggregating $i$ and $j$ is the difference between their individual contributions and the aggregated contribution of $i\cup j$:
	\begin{align*}
		d(i,j) = I_{i}(X,Y) + I_{j}(X,Y) - I_{i \cup j}(X,Y).  
	\end{align*}
	The function $d$ may be regarded as a similarity measure. It satisfies $d(i,j) = 0$ if and only if $p(Y|X = i) = p(Y|X = j)$; i.e. the demographic profiles of locations $i$ and $j$ are identical. It is also symmetric in $i$ and $j$. We can use it to motivate a form of agglomerative hierarchical clustering, where at each phase of the algorithm we compute 
	\begin{equation}
		(i^*, j^*) = \argmin_{i \text{ neighbors } j} \; d(i,j)\; \label{eq:cluster_opt}
	\end{equation}
	and combine $i^*$ and $j^*$ into a single new locale, repeating until only a specified number $n$ locales remain. 
	Equation \eqref{eq:cluster_opt} defines \emph{spatially constrained, information-maximizing agglomerative clustering}. 
	Since the algorithm is greedy, it possesses no guarantees of optimality, but in practice its performance leads to intuitive, demographically-coherent regions. 
	The resulting regions may be viewed as coarse-grainings of the spatial structure, which emphasize high-level patterns and abstract away details.

	Figure \ref{fig:clusterings} demonstrates this information-maximizing agglomerative clustering for Atlanta and Philadelphia, using 4 and 20 clusters. In Atlanta, the algorithm immediately highlights the prominent North-South divide, while in Philadelphia it segments two predominantly Black regions on either side of the Schuykill river from the rest of the city. As we allow the algorithm to use progressively more clusters, it highlights progressively more subtle patterns of variation. Strikingly, in Atlanta just 4 clusters are sufficient to capture 0.27 nats of information, where in Philadelphia 20 clusters are needed. Philadelphia has more complex sociospatial trends: compared to Atlanta, more complex models are necessary to model it to a similar degree of accuracy.

	\begin{figure}
		\centering
			% \begin{subfigure}[b]{0.45\linewidth}
			\includegraphics[width = .5\linewidth]{figs/example_cluster_map.pdf}
			% \caption{} \label{subfig:detroit_map}
			% \end{subfigure}
			% \begin{subfigure}[b]{0.45\linewidth}
			\includegraphics[width = .45\linewidth]{figs/example_clusters_detailed.pdf}
			% \caption{} \label{subfig:detroit_groups}
			% \end{subfigure}
			\caption{
				Multigroup sociospatial structure in Detroit. 
				\emph{(Left)}: Map of six clusters. 
				\emph{(Right)}: Joint distribution of demographics and clusters. 
				Shadings reflect population proportions within clusters. 
				Numerical labels give raw population proportions; for example, 29\% of the population is Black and resides in cluster B. 
				Detroit has $H(Y) = 1.12$, $I(X,Y) = 0.57$, and $J(X,Y) = 0.08$. 
				This clustering captures $I(C_6,Y) = 0.36$ nats of information.
			} \label{fig:detroit}
	\end{figure}

	Information-maximizing agglomerative clustering also applies to multigroup analysis.  Figure \ref{fig:detroit} shows a multigroup clustering in Detroit and the surrounding Wayne County. The prominent clusters A and B reflect the primary white/black division in the city, with A representing the white suburbs and B the denser black inner city. Cluster C is interpretable as a transitional area between these two populations, with both white and black residents represented. Cluster D coincides with the historically distinct, predominantly white town of Grosse Pointe. Cluster E coincides with the city of Hamtramck, whose history has been shaped by high levels of immigration from Europe and Asia. Finally, Cluster F coincides with the Hispanic neighborhood Mexicantown.

	This technique also enables novel quantitative methods for studying traditional concerns in segregation studies. According to \cite{Massey1988}, ``\emph{concentration} refers to the degree of a group's agglomeration in urban space.'' The concept of concentration invites us to consider the composition and density of each cluster. For example, whites make up 70\% of ``their'' cluster A, while black residents make up 86\% of ``their'' cluster B. Cluster B is also more densely populated than Cluster A: it contains 34\% of Wayne County's population, but just 19\% of the analyzed geographic area, making it more than twice as dense as Cluster A. Thus, in and around Detroit, black neighborhoods tend to be both more racially homogeneous and more densely packed in urban space than white ones. We can also learn about pairwise group exposure. For example, from Figure  \ref{fig:detroit}, Asian residents tend to be found alongside white residents, but more rarely alongside black or Hispanic residents. Hispanic residents tend to group in Mexicantown and in predominantly white neighborhoods, but not black ones.

\subsection*{Local Information and Spatial Complexity}
	The mean local information measures the number and character of spatial transitions between demographically distinct regions. When there are many, we would expect that information-maximizing agglomerative clustering would require more clusters to capture similar amounts of information. We can empirically quantify the relationship between $J(X,Y)$ and the performance of information-maximizing clustering as follows. Consider the between-groups information $I(C_n,Y)$ plotted as a function of the number $n$ of clusters. The result is a curve reflecting how information scales at varying levels of aggregation, as shown in Figure \ref{fig:AUC} shows examples of these curves for Atlanta, Detroit, and Philadelphia.
